Simple repression draft

### 2020-10-22 doc/resubmission/appendix_01_nonbursty.tex
final pass before resubmission

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for both thermodynamic and kinetic promoter models
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these models, the natural mathematicization of their
cartoons is either an equilibrium model based on statisticla mechanics, or a
chemical master equation. These derivations will go through the specifics of the
models in Figure~\ref{fig1:means_cartoons}. We point the readers to some other
excellent reviews on the general
frameworks~\cite{Bintu2005a,Sanchez2013,Phillips2015a}.

\subsection{Thremodynamic models of gene regulation}
The first class of models we will explore are the so-called thermodynamic models
of gene regulation~\cite{Ackers1982}. The premise for these models is that we
imagine the steps leading to transcription --binding and unbinding of
transcription factors and of RNAP-- to occur in a timescale much faster than the
downstream processes --open complex formation and RNAP
escape~\cite{Browning2004}. What this allows us to do is to assume that the
steps leading to transcription can be modeled as being in quasi-equilibrium.
This is enormously advantageous since we can use the theoretical tools of 
equilibrium statistical mechanics to model such process.

@@

% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
% intuitive analog to model 1, with just two states, repressor bound or unbound,
% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
% the unbound state, single mRNA transcripts are produced as a Poisson process
% with some characteristic rate $r$. The current model by contrast produces, at
% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
% The burst sizes are assumed to be geometrically distributed with
% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
% when we derive this model as a certain limiting case of model 4.

% From this intuitive picture and by analogy to model 1, then, it should be
% plausible that the mean mRNA level is
% \begin{equation}
% \langle m\rangle = \frac{k_i b}{\gamma}
%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
% \end{equation}
% which will turn out to be correct from a careful calculation. For now, we simply
% note that just like model 1, the fold-change becomes
% \begin{equation}
% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
% \end{equation}
% with $\rho=1$ also like model 1.
% We will also see later how this model emerges as a natural limit of model 4.
% \mrm{ENDS HERE}

% \mrm{THIS NEEDS TO BE FIXED}
% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}

% Our second model of constitutive transcription posits an architecture in which
% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
% transcription initiation results in a state transition from the bound to the
% unbound state, reflecting the microscopic reality that an RNAP that has begun to
% elongate a transcript is no longer available at the start site to begin another.
% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
% given by
% \begin{align}
%     \nu = 1 -
%         \frac{r\rate{k}{P}{+}}
%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
% \label{eq:model2_fano}
% \end{align}
% The problem with this picture is that the Fano factor is always $<1$. To make
%  contact with the experimental reality of $\nu>1$ as shown in
%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
%  While this model adds an appealing element of microscopic reality, we are
%  forced to reject it as the additional complexity is unable to capture the
%  phenomenology of interest. Obviously the promoter state does in fact proceed
%  through cycles of RNAP binding, initiating, and elongating, but it seems that
%  the super-Poissonian noise in mRNA copy number we want to model must be
%  governed by other features of the system.

% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
% Initiation and Escape.}

% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
% that once RNAP initially binds the promoter region, a multitude of distinct
% steps occur sequentially before RNAP finally escapes into the elongation phase.
% Perhaps adding some of this mechanistic detail as shown in model 3 of
% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
% simplest refinement of that model could consider open complex formation and
% promoter escape as separate steps rather than as a single effective step. In
% other words, we construct model 3 by adding a single extra state to model 2, and
% we will label the two RNAP-bound states as the closed and open complexes,
% despite the true biochemical details certainly being more complex. For example,
% earlier work extended this model by adding an additional repressor bound state
% and did not explicitly consider the limit with no repressor that we analyze
% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
% all the relevant biochemical detail; this is an exploratory search for the
% important features that a model needs to include to square with the known
% experimental reality of constitutive expression.

% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
% has Fano factor $\nu<1$. We again leave the algebraic details for
% Appendix~\ref{sec:non_bursty} and merely state the result that
% \begin{align}
% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
% \label{eq:model3_fano}
% \end{align}
% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
% tidiness. This is necessarily less than 1 for arbitrary rate constants.

% In fact, we suspect \textit{any} model in which transcription proceeds through a
% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
% the waiting time distribution to traverse the cycle with the waiting time for a
% Poisson promoter (model 1) with the same mean time. The latter is simply an
% exponential distribution. The former is a convolution of multiple exponentials,
% and intuitively the waiting time distribution for a multistep process
% should be more peaked with a smaller fractional width than a single
% exponential with the same mean.
% A less disperse waiting time distribution means transcription initations are
% more uniformly distributed in time relative to a Poisson process. Hence the
% distribution of mRNA over a population of cells should be less variable compared
% to Poisson, giving $\nu<1$.
% (In Appendix~\ref{sec:non_bursty} we present a more precise version
% of the intuitive arguments in this paragraph.)
% Regardless of the merits of this model in describing the noise properties of
% constitutive transcription initiation, it ultimately fails the simplest
% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
% we must discard this mechanistic picture and search elsewhere.
% \mrm{ENDS HERE}

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation. For model 1 we have the master equation
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for kinetic model 2 in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to Eq.~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model 2 can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

\subsubsection{Summary of general results}
For ease of reference, we collect and reprint here the key results derived in
this section that are used in the main text and subsequent subsections. Mean
mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx}
and \ref{eq:generic_fano_appdx}, which are
\begin{equation}
\langle{m}\rangle
= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
\label{eq:mean_m_appdx_ref}
\end{equation}
and
\begin{equation}
\nu = 1 - \langle{m}\rangle
+ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
        {\gamma \langle{m}\rangle},
\label{eq:fano_appdx_ref}
\end{equation}
respectively. To compute these two quantities, we need the expressions for
$\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
Eqs.~\ref{eq:generic_cme_vecm0} and \ref{eq:generic_cme_deriv_040},
respectively, which are
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
\label{eq:vecm0_appdx_ref}
\end{equation}
and
\begin{equation}
(\mathbf{K} - \mathbf{R_D}
+ \mathbf{R_A} - \gamma\mathbf{I}) \langle\vec{m}\rangle
= - \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:vecm_appdx_ref}
\end{equation}
Some comments are in order before we consider particular models. First, note
that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother solving for
all components of the vectors $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$, but only the components which are multiplied by nonzero
elements of $\mathbf{R_A}$. The only component of $\langle\vec{m}^0\rangle$ that
ever survives is the transciptionally active state, and for the models we
consider here, there is only ever one such state. This will save us some amount
of algebra below.

@@

This first model (model 1 in Figure~\ref{fig1:means_cartoons}) is the one that
we used in the previous section to motivate the derivation of the chemical
master equation. Eq.~\ref{eq:poisson_promoter_cme_appdx} shows the two coupled
master equations, but for completeness we reprint it again as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\end{split}
\label{eq:model1_kinetic_cme}
\end{align}
This is a direct transcription of the states and rates in
Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\end{equation}
where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript and
denote both simply by $\mathbf{R}$. Since our interest is only in steady-state
we dropped the time dependence as well.

@@

The master equation governing this model is analogous to
Eq.~\ref{eq:model1_kinetic_cme} for model 1 above. The main subtlety arises
since transcription initiation accompanies a promoter state change. The full
master equation for model 2 in Figure~\ref{fig1:means_cartoons} is then of the
form
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
    \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U},
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\end{split}
\end{align}

@@

\subsubsection{Mean mRNA}
As for model 1, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being found
in repressor bound, empty, or RNAP-bound states, respectively. Only $p_P$ is
necessary to find since
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 \\ r p_P \\ 0 \end{pmatrix}.
\label{eq:model2_deriv_appdx_020}
\end{equation}
Then Eq.~\ref{eq:vecm0_appdx_ref} for $\langle\vect{m}\rangle$ reads
\begin{equation}
\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
        k_R^- & -k_R^+ -k_P^+ & k_P^- + r\\
        0 & k_P^+ & -k_P^- - r
\end{pmatrix}
\begin{pmatrix}
    p_R \\ p_E \\ p_P
\end{pmatrix}
= 0.
\label{eq:model2_K-R+R_for_m0}
\end{equation}
Discarding the middle row as redundant and incorporating the normalization
condition leads to a set of three linearly independent equations, namely
\begin{align}
-k_R^- p_R + k_R^+ p_E &= 0 \\
k_P^+ p_E + (-k_P^- - r) p_P &= 0 \\
p_R + p_E + p_P &= 1.
\end{align}
Using $p_R = 1 - p_E - p_P$ to eliminate $p_R$ in the first and solving the
resulting equation for $p_E$ gives
$p_E = (1 - p_P){k_R^-}/{(k_R^- + k_R^+)}$.
Substituting this for $p_E$ in the second equation gives an equation in
$p_P$ alone which is
\begin{equation}
k_P^+ k_R^- (1-p_P) - (k_P^- + r)(k_R^+ + k_R^-) p_P = 0
\end{equation}
and solving for $p_P$ gives
\begin{equation}
p_P = \frac{k_P^+ k_R^-}{k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
\end{equation}
Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by
$\mathbf{R_A}$ produces
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= r \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}
\begin{pmatrix} 0 \\ 1 \\ 0 \end{pmatrix}
\end{equation}
from which $\langle{m}\rangle$ follows readily,
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma}
        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
\label{eq:model2_meanm_appdx}
\end{equation}

>>>>

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for both thermodynamic and kinetic promoter models
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these models, the natural mathematicization of their
cartoons is either an equilibrium model based on statistical mechanics, or a
chemical master equation. These derivations will go through the specifics of the
models in Figure~\ref{fig1:means_cartoons}. We point the readers to some other
excellent reviews on the general
frameworks~\cite{Bintu2005a,Sanchez2013,Phillips2015a}.

\subsection{Thermodynamic models of gene regulation}
The first class of models we will explore are the so-called thermodynamic models
of gene regulation~\cite{Ackers1982}. The premise for these models is that we
imagine the steps leading to transcription --binding and unbinding of
transcription factors and of RNAP-- to occur in a timescale much faster than the
downstream processes --open complex formation and RNAP
escape~\cite{Browning2004}. What this allows us to do is to assume that the
steps leading to transcription can be modeled as being in quasi-equilibrium.
This is enormously advantageous since we can use the theoretical tools of
equilibrium statistical mechanics to model such process.

@@

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation. For model 1 we have the master equation
\begin{equation}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{equation}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for kinetic model 2 in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to Eq.~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model 2 can be written
\begin{equation}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{equation}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

\subsubsection{Summary of general results}
For ease of reference, we collect and reprint here the key results derived in
this section that are used in the main text and subsequent subsections. Mean
mRNA copy number and Fano factor are given by Eqs.~\ref{eq:generic_mean_m_appdx}
and \ref{eq:generic_fano_appdx}, which are
\begin{equation}
\langle{m}\rangle
= \frac{1}{\gamma} \vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
\label{eq:mean_m_appdx_ref}
\end{equation}
and
\begin{equation}
\nu = 1 - \langle{m}\rangle
+ \frac{\vec{1}^\dagger\cdot \mathbf{R_A}\langle\vec{m}\rangle}
        {\gamma \langle{m}\rangle},
\label{eq:fano_appdx_ref}
\end{equation}
respectively. To compute these two quantities, we need the expressions for
$\langle\vec{m}^0\rangle$ and $\langle\vec{m}\rangle$ given by solving
Eqs.~\ref{eq:generic_cme_vecm0} and \ref{eq:generic_cme_deriv_040},
respectively, which are
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle = 0
\label{eq:vecm0_appdx_ref}
\end{equation}
and
\begin{equation}
(\mathbf{K} - \mathbf{R_D}
+ \mathbf{R_A} - \gamma\mathbf{I}) \langle\vec{m}\rangle
= - \mathbf{R_A} \langle\vec{m}^0\rangle.
\label{eq:vecm_appdx_ref}
\end{equation}
Some comments are in order before we consider particular models. First, note
that to obtain $\langle\vec{m}\rangle$ and $\nu$, we need not bother solving for
all components of the vectors $\langle\vec{m}^0\rangle$ and
$\langle\vec{m}\rangle$, but only the components which are multiplied by nonzero
elements of $\mathbf{R_A}$. The only component of $\langle\vec{m}^0\rangle$ that
ever survives is the transcriptionally active state, and for the models we
consider here, there is only ever one such state. This will save us some amount
of algebra below.

@@

This first model (model 1 in Figure~\ref{fig1:means_cartoons}) is the one that
we used in the previous section to motivate the derivation of the chemical
master equation. Eq.~\ref{eq:poisson_promoter_cme_appdx} shows the two coupled
master equations, but for completeness we reprint it again as
\begin{equation}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\end{split}
\label{eq:model1_kinetic_cme}
\end{equation}
This is a direct transcription of the states and rates in
Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\end{equation}
where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript and
denote both simply by $\mathbf{R}$. Since our interest is only in steady-state
we dropped the time dependence as well.

@@

The master equation governing this model is analogous to
Eq.~\ref{eq:model1_kinetic_cme} for model 1 above. The main subtlety arises
since transcription initiation accompanies a promoter state change. The full
master equation for model 2 in Figure~\ref{fig1:means_cartoons} is then of the
form
\begin{equation}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
    \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U},
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\end{split}
\end{equation}

@@

\subsubsection{Mean mRNA}
As for model 1, we must first find $\mathbf{R_A} \langle\vect{m}^0\rangle$.
Denote its components as $p_R$, $p_E$, $p_P$, the probabilities of being found
in repressor bound, empty, or RNAP-bound states, respectively. Only $p_P$ is
necessary to find since
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 \\ r p_P \\ 0 \end{pmatrix}.
\label{eq:model2_deriv_appdx_020}
\end{equation}
Then Eq.~\ref{eq:vecm0_appdx_ref} for $\langle\vect{m}\rangle$ reads
\begin{equation}
\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
        k_R^- & -k_R^+ -k_P^+ & k_P^- + r\\
        0 & k_P^+ & -k_P^- - r
\end{pmatrix}
\begin{pmatrix}
    p_R \\ p_E \\ p_P
\end{pmatrix}
= 0.
\label{eq:model2_K-R+R_for_m0}
\end{equation}
Discarding the middle row as redundant and incorporating the normalization
condition leads to a set of three linearly independent equations, namely
\begin{eqnarray}
-k_R^- p_R + k_R^+ p_E &= 0 \\
k_P^+ p_E + (-k_P^- - r) p_P &= 0 \\
p_R + p_E + p_P &= 1.
\end{eqnarray}
Using $p_R = 1 - p_E - p_P$ to eliminate $p_R$ in the first and solving the
resulting equation for $p_E$ gives
$p_E = (1 - p_P){k_R^-}/{(k_R^- + k_R^+)}$.
Substituting this for $p_E$ in the second equation gives an equation in
$p_P$ alone which is
\begin{equation}
k_P^+ k_R^- (1-p_P) - (k_P^- + r)(k_R^+ + k_R^-) p_P = 0
\end{equation}
and solving for $p_P$ gives
\begin{equation}
p_P = \frac{k_P^+ k_R^-}{k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
\end{equation}
Substituting this in Eq.~\ref{eq:model2_deriv_appdx_020} and multiplying by
$\mathbf{R_A}$ produces
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= r \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}
\begin{pmatrix} 0 \\ 1 \\ 0 \end{pmatrix}
\end{equation}
from which $\langle{m}\rangle$ follows readily,
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma}
        \frac{k_P^+ k_R^-} {k_P^+ k_R^- + (k_P^- + r)(k_R^+ + k_R^-)}.
\label{eq:model2_meanm_appdx}
\end{equation}
